---
title: "FinalProject"
author: "Branson Vu, Blake Tran, Daylan Petershagen, Emmett Schoenhals"
format: pdf
editor: visual
---

# Exploratory Data Analysis

```{r}
#| label: Setup
#| message: false
#| warning: false

# libraries
library(dplyr)
library(ggplot2)
library(corrplot)
# GGally is a package that creates nice looking scatterplot matrice
library(GGally)
library(tidyr)
#library(car)

# Load the RData file, no need to directly click on the file to load it
load("concrete_data.RData")
```

## Data Overview

```{r}
#| label: Data-Overview

# Dimensions and types
str(concrete_train)

# Summary stats for all variables
summary(concrete_train)
```

```{r}
#| label: Missing-Values-and-Proportion-of-Zeros

# Check for missing values in train set
colSums(is.na(concrete_train))

# Proportion of zeros per covariate (slag, ash, cemet_superplasticizer should have many)
round(colMeans(concrete_train == 0), 3)
```

## Response Variable: ccs

```{r}
#| label: Shapiro-Tests

# normality tests on raw and log-transformed ccs
shapiro.test(concrete_train$ccs)
shapiro.test(log(concrete_train$ccs))
```

Notes: Based on the Shapiro-Wilk tests and skewness checks, we will use the raw ccs values for modeling because the log transformation overcorrects the data and actually introduces a more severe left-skew.

```{r}
#| label: Response-Distribution
#| fig-width: 8
#| fig-height: 4

# Histogram + density of ccs (concrete compressive strength)
ggplot(concrete_train, aes(x = ccs)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  labs(title = "Distribution of CCS", x = "Compressive Strength (MPa)", y = "Density") +
  theme_minimal()
```

## Covariate Distributions

```{r}
#| label: Covariate-Histograms
#| fig-width: 10
#| fig-height: 8

# Histograms of all 8 covariates
concrete_train %>%
  select(-ccs) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(title = "Covariate Distributions") +
  theme_minimal()
```

## Correlations

```{r}
#| label: Correlation-Matrix
#| fig-width: 8
#| fig-height: 7

# pearson correlation matrix
cor_mat <- cor(concrete_train)
corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0,0,1,0))
```

```{r}
#| label: Correlations-with-Response

# Correlations with ccs sorted by absolute value
cors <- cor(concrete_train)[, "ccs"]
sort(abs(cors[-length(cors)]), decreasing = TRUE)
```

Notes: `cement` ranks #1 with a correlation of \~0.50. Age is the next highest positive correlation. These covariates will be used for the model.

## Scatterplots vs Response

```{r}
#| label: Scatter-vs-ccs
#| fig-width: 10
#| fig-height: 8

# Scatterplots of each covariate against ccs
concrete_train %>%
  pivot_longer(-ccs, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, y = ccs)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.7) +
  facet_wrap(~variable, scales = "free_x", ncol = 3) +
  labs(title = "Covariates vs CCS", y = "Compressive Strength (MPa)") +
  theme_minimal()
```

Notes: Most covariates show roughly linear relationships with ccs, but `age` stands out — the loess smoother reveals a steep rise in strength during early curing (days 1–28) that flattens at later ages. This diminishing-returns pattern suggests a `log(age)` transformation may better capture the relationship than a raw linear term.

## Multicollinearity Check

```{r}
#| label: Pairs-Top-Covariates
#| fig-width: 10
#| fig-height: 10
#| message: false

# Pairwise plot of top correlated covariates + response
top_vars <- c("cement", "age", "water", "superplasticizer", "ccs")
ggpairs(concrete_train[, top_vars],
        lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5))) +
  theme_minimal()
```

## Frequentist Screening

```{r}
freq_fit <- lm(ccs ~ ., data = concrete_train)
summary(freq_fit)
# check multicollinearity
#car::vif(freq_fit)
```

Age, Cement, Slag, Ash, Water are likely significant covariates, but we include all in the model and compare via LOO.

## Modeling Assumptions

Before fitting our Bayesian linear regression, we state the key assumptions underlying the model:

1.  **Linearity**: The response (ccs) is a linear function of the covariates. The scatterplots with loess smoothers above show roughly linear trends for the strongest predictors (cement, water, age), supporting this assumption.
2.  **Independence**: Each observation represents a separate concrete mixture, so we assume independence across samples.
3.  **Normality of errors**: The residuals are assumed to follow a Normal distribution. The Shapiro-Wilk test on raw ccs showed mild deviation (common with N=800), but the distribution is reasonably symmetric with a skewness of only +0.41.
4.  **Constant variance (homoscedasticity)**: The variance of residuals is assumed constant across all fitted values. We chose raw ccs over log(ccs) based on the Shapiro tests; residual plots after fitting will verify this.
5.  **No perfect multicollinearity**: VIF values from the frequentist screening are elevated (5-8 range) for several covariates due to the mixture nature of concrete data, but none are extreme. Bayesian priors provide regularization that mitigates multicollinearity.

## Proposed Model

We fit a **Bayesian linear regression** with the following specification:

**Likelihood:** $$y_i \mid \boldsymbol{\beta}, \sigma^2 \sim N(\alpha + \mathbf{x}_i^\top \boldsymbol{\beta},\; \sigma^2)$$

Where $y_i$ is the raw compressive strength (ccs) and $\mathbf{x}_i$ is the standardized covariate vector for observation $i$.

**Priors:**

-   Intercept: $\alpha \sim N(0, 10)$
-   Slopes: $\beta_j \sim N(0, 5)$ for $j = 1, \ldots, 8$ (weakly informative on standardized scale)
-   Residual SD: $\sigma \sim \text{Half-Cauchy}(0, 5)$

**Prior sensitivity plan:** We will refit the model under tight priors ($\beta \sim N(0,1)$) and wide priors ($\beta \sim N(0,50)$), then compare posterior means and 95% credible intervals across all three specifications. If the posteriors are stable, our inference is data-driven. We will also test sensitivity to model formulation by adding a cement $\times$ age interaction term and a log(age) transformation, comparing all variants via LOO-CV.

# Model Fit and Assessment

```{r}
#| label: Baseline-Stan
#| message: false
#| warning: false

library(rstan)
#library(rstanarm)
library(loo)
# standardize covariates
X_raw <- concrete_train %>% select(-ccs)
X_means <- colMeans(X_raw)
X_sds   <- apply(X_raw, 2, sd)
X_scaled <- scale(X_raw)
```

```{r}
stan_concrete_model <- "
data {
  int<lower=0> N;          // number of observations
  int<lower=0> K;          // number of covariates
  matrix[N, K] X;          // covariate matrix (standardized)
  vector[N] y;             // response (raw ccs)
}

parameters {
  real alpha;              // intercept
  vector[K] beta;          // slopes
  real<lower=0> sigma;     // residual sd
}

model {
  // priors
  alpha ~ normal(0, 10);
  beta  ~ normal(0, 5);
  sigma ~ cauchy(0, 5);

  // likelihood
  y ~ normal(alpha + X * beta, sigma);
}

generated quantities {
  vector[N] log_lik;       // for LOO-CV
  for (n in 1:N)
    log_lik[n] = normal_lpdf(y[n] | alpha + X[n] * beta, sigma);
}
"
  
```

```{r}
#| label: Fit-Base-Model

# prepare data list for Stan (using raw ccs based on EDA — log overcorrects skew)
stan_data <- list(
  N = nrow(concrete_train),
  K = ncol(X_scaled),
  X = X_scaled,
  y = concrete_train$ccs
)

fit <- stan(model_code = stan_concrete_model, data = stan_data,
            chains = 4, iter = 2000, warmup = 1000, seed = 40233)
```

## Convergence Diagnostics

We check convergence to ensure MCMC chains have mixed well and our posterior samples are reliable.

```{r}
#| label: Model-Summary

# posterior summary — check R-hat (should be < 1.01) and n_eff
print(fit, pars = c("alpha", "beta", "sigma"))
```

```{r}
#| label: Traceplots
#| fig-width: 10
#| fig-height: 8

library(bayesplot)

# traceplots — chains should look like "hairy caterpillars" (well-mixed)
posterior <- as.array(fit)
mcmc_trace(posterior, pars = c("alpha", paste0("beta[", 1:8, "]"), "sigma"))
```

### Gelman-Rubin (R-hat)

R-hat compares between-chain and within-chain variance. Values near 1.0 indicate convergence; values \> 1.01 suggest chains haven't converged.

```{r}
#| label: Gelman-Rubin

# extract R-hat for all parameters
fit_summary <- summary(fit)$summary
rhat_vals <- fit_summary[, "Rhat"]

# show R-hat for model parameters only
param_names <- c("alpha", paste0("beta[", 1:8, "]"), "sigma")
rhat_params <- rhat_vals[param_names]
print(round(rhat_params, 4))

# flag any R-hat > 1.01
cat("\nAny R-hat > 1.01?", any(rhat_params > 1.01, na.rm = TRUE), "\n")
```

### Geweke Diagnostic

Geweke tests whether the mean of the first 10% and last 50% of the chain are equal. Z-scores outside (-2, 2) suggest non-convergence.

```{r}
#| label: Geweke-Diagnostic

# bayesian analysis, specifically MCMC sims
library(coda)

# convert to coda mcmc.list for Geweke test
posterior_list <- As.mcmc.list(fit)

# Geweke z-scores for each chain (first chain shown)
geweke_results <- geweke.diag(posterior_list[[1]])
print(geweke_results)
```

### Effective Sample Size (ESS)

ESS estimates the number of independent samples from the posterior. We want ESS \> 1000 for all parameters to ensure stable posterior estimates.

```{r}
#| label: ESS-Check

# extract ESS (n_eff) for model parameters
ess_vals <- fit_summary[param_names, "n_eff"]
print(round(ess_vals))

# flag any ESS < 1000
cat("\nAny ESS < 1000?", any(ess_vals < 1000, na.rm = TRUE), "\n")
```

## Prior Sensitivity Analysis

To ensure these estimates are trustworthy and not artifacts of our prior choice, we refit the model under three prior specifications:

-   **Base model**: `beta ~ N(0, 5)` — weakly informative, allows moderate coefficient values
-   **Tight priors**: `beta ~ N(0, 1)` — more regularizing, shrinks coefficients toward zero
-   **Wide priors**: `beta ~ N(0, 50)` — essentially flat/uninformative, lets data fully determine estimates

If the posterior means and credible intervals are stable across all three, our results are data-driven and robust.

```{r}
#| label: Prior-Sensitivity-Models

# Tight priors: stronger regularization, pulls coefficients toward 0
# If posteriors shift significantly vs base → prior is too influential
stan_model_tight <- "
data {
  int<lower=0> N;
  int<lower=0> K;
  matrix[N, K] X;
  vector[N] y;
}
parameters {
  real alpha;
  vector[K] beta;
  real<lower=0> sigma;
}
model {
  alpha ~ normal(0, 2);
  beta  ~ normal(0, 1);
  sigma ~ cauchy(0, 2);
  y ~ normal(alpha + X * beta, sigma);
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N)
    log_lik[n] = normal_lpdf(y[n] | alpha + X[n] * beta, sigma);
}
"

# Wide priors: nearly flat, lets data speak entirely for itself
# If posteriors match the base model, then the data is informative enough to overcome any prior
stan_model_wide <- "
data {
  int<lower=0> N;
  int<lower=0> K;
  matrix[N, K] X;
  vector[N] y;
}
parameters {
  real alpha;
  vector[K] beta;
  real<lower=0> sigma;
}
model {
  alpha ~ normal(0, 100);
  beta  ~ normal(0, 50);
  sigma ~ cauchy(0, 25);
  y ~ normal(alpha + X * beta, sigma);
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N)
    log_lik[n] = normal_lpdf(y[n] | alpha + X[n] * beta, sigma);
}
"

fit_tight <- stan(model_code = stan_model_tight, data = stan_data,
                  chains = 4, iter = 2000, warmup = 1000, seed = 40233)

fit_wide  <- stan(model_code = stan_model_wide, data = stan_data,
                  chains = 4, iter = 2000, warmup = 1000, seed = 40233)
```

```{r}
#| label: Prior-Sensitivity-Comparison
#| fig-width: 10
#| fig-height: 6

# compare posterior means across all three prior settings
# if values are nearly identical, then our posterior means are robust
base_means  <- summary(fit)$summary[param_names, "mean"]
tight_means <- summary(fit_tight)$summary[param_names, "mean"]
wide_means  <- summary(fit_wide)$summary[param_names, "mean"]

comparison <- data.frame(
  Parameter = param_names,
  Base  = round(base_means, 3),
  Tight = round(tight_means, 3),
  Wide  = round(wide_means, 3)
)
print(comparison)

# also compare 95% credible interval widths for each prior setting
cat("\n=== 95% Credible Intervals: Base Model ===\n")
print(round(summary(fit)$summary[param_names, c("2.5%", "97.5%")], 3))

cat("\n=== 95% Credible Intervals: Tight Priors ===\n")
print(round(summary(fit_tight)$summary[param_names, c("2.5%", "97.5%")], 3))

cat("\n=== 95% Credible Intervals: Wide Priors ===\n")
print(round(summary(fit_wide)$summary[param_names, c("2.5%", "97.5%")], 3))
```

If the credible intervals are similar across all three prior specifications, then our inference is stable and we can confidently report the base model's posterior means and 95% intervals as our final estimates.

```{r}
#| label: Prior-Sensitivity-Density
#| fig-width: 10
#| fig-height: 6

# visual comparison of posterior densities across prior choices
post_base  <- as.matrix(fit)[, paste0("beta[", 1:8, "]")]
post_tight <- as.matrix(fit_tight)[, paste0("beta[", 1:8, "]")]
post_wide  <- as.matrix(fit_wide)[, paste0("beta[", 1:8, "]")]

# overlay densities for cement (strongest predictor)
# densities that overlap = prior doesn't matter = trustworthy estimate
plot_df <- data.frame(
  value = c(post_base[,1], post_tight[,1], post_wide[,1]),
  Prior = rep(c("Base N(0,5)", "Tight N(0,1)", "Wide N(0,50)"), each = nrow(post_base))
)

ggplot(plot_df, aes(x = value, fill = Prior)) +
  geom_density(alpha = 0.4) +
  labs(title = "Prior Sensitivity: Posterior of cement coefficient",
       subtitle = "Overlapping densities indicate data-driven posterior",
       x = "beta[1] (cement)") +
  theme_minimal()
```

## LOO-CV Comparison

LOO-CV (Leave-One-Out Cross-Validation) estimates out-of-sample predictive performance. Higher ELPD = better model. We compare the base model across prior specifications.

```{r}
#| label: Loo-Comparison

# LOO for each prior setting
loo_base  <- loo(extract_log_lik(fit, merge_chains = FALSE))
loo_tight <- loo(extract_log_lik(fit_tight, merge_chains = FALSE))
loo_wide  <- loo(extract_log_lik(fit_wide, merge_chains = FALSE))

# compare
loo_compare(loo_base, loo_tight, loo_wide)
```

Note: Base Model and Model with Wide Priors yield same results

## Model Formulation Sensitivity

The EDA scatterplots revealed that `age` has a nonlinear (diminishing-returns) relationship with ccs, while `cement` and `age` are the two strongest predictors. We test two alternative formulations: (1) adding a `cement * age` interaction term, and (2) replacing `age` with `log(age)` to capture the curing curve. LOO-CV determines which formulation best improves out-of-sample predictions.

```{r}
#| label: Model-Interaction
#| message: false

# Model 2: add cement*age interaction (two strongest predictors) as beta[9]
X_interact <- cbind(X_scaled, cement_age = X_scaled[,"cement"] * X_scaled[,"age"])

stan_data_interact <- list(
  N = nrow(concrete_train),
  K = ncol(X_interact),
  X = X_interact,
  y = concrete_train$ccs
)

fit_interact <- stan(model_code = stan_concrete_model, data = stan_data_interact,
                     chains = 4, iter = 2000, warmup = 1000, seed = 40233)
```

```{r}
#| label: Model-Log-Age
#| message: false

# Model 3: use log(age) instead of raw age (diminishing returns pattern in EDA)
X_log_age <- X_raw
X_log_age$age <- log(X_log_age$age)
X_log_age_scaled <- scale(X_log_age)

stan_data_log_age <- list(
  N = nrow(concrete_train),
  K = ncol(X_log_age_scaled),
  X = X_log_age_scaled,
  y = concrete_train$ccs
)

fit_log_age <- stan(model_code = stan_concrete_model, data = stan_data_log_age,
                    chains = 4, iter = 2000, warmup = 1000, seed = 40233)
```

```{r}
#| label: Formulation-Loo-Comparison

# LOO comparison of model formulations
loo_interact <- loo(extract_log_lik(fit_interact, merge_chains = FALSE))
loo_log_age  <- loo(extract_log_lik(fit_log_age, merge_chains = FALSE))

cat("=== Model Formulation Comparison ===\n")
loo_compare(loo_base, loo_interact, loo_log_age)
```

Note: `fit_log_age` is the better model compared to the base model and the model with the cement\*age covariate.

```{r}
#| label: Formulation-Summary

# summary of interaction model
cat("=== Interaction Model (cement*age) ===\n")
print(fit_interact, pars = c("alpha", paste0("beta[", 1:9, "]"), "sigma"))

# summary of log-age model
cat("\n=== Log-Age Model ===\n")
print(fit_log_age, pars = c("alpha", paste0("beta[", 1:8, "]"), "sigma"))
```

# Interpretation of Results

## Covariate Significance Summary

```{r}
#| label: Covariate-Significance

# extract posterior samples for all betas from the best model (fit_log_age)
beta_samples <- as.matrix(fit_log_age)[, paste0("beta[", 1:8, "]")]
covariate_names <- colnames(X_log_age_scaled)

# build summary table
sig_table <- data.frame(
  Covariate = covariate_names,
  
  # posterior mean
  Post_Mean = round(colMeans(beta_samples), 3),
  
  # 95% credible interval bounds
  CI_Lower = round(apply(beta_samples, 2, quantile, 0.025), 3),
  CI_Upper = round(apply(beta_samples, 2, quantile, 0.975), 3),
  
  # probability of direction: max(P(beta > 0), P(beta < 0))
  # values near 1.0 = strong directional evidence
  Prob_Direction = round(apply(beta_samples, 2, function(x) max(mean(x > 0), mean(x < 0))), 3),
  
  # does the 95% CI exclude zero?
  Significant = apply(beta_samples, 2, function(x) {
    ci <- quantile(x, c(0.025, 0.975))
    ifelse(ci[1] > 0 | ci[2] < 0, "YES", "NO")
  })
)

rownames(sig_table) <- NULL
print(sig_table)

# quick summary
cat("\n--- Significant covariates (95% CI excludes 0) ---\n")
cat(paste(sig_table$Covariate[sig_table$Significant == "YES"], collapse = ", "), "\n")

cat("\n--- Non-significant covariates (95% CI includes 0) ---\n")
cat(paste(sig_table$Covariate[sig_table$Significant == "NO"], collapse = ", "), "\n")
```

```{r}
#| label: Significance-Plot
#| fig-width: 9
#| fig-height: 5

# visual: color-coded credible intervals (green = significant, red = not)
sig_table$color <- ifelse(sig_table$Significant == "YES", "Significant", "Not Significant")

ggplot(sig_table, aes(x = Post_Mean, y = reorder(Covariate, abs(Post_Mean)), color = color)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Significant" = "green", "Not Significant" = "red")) +
  labs(title = "Covariate Significance: 95% Credible Intervals",
       subtitle = "Green intervals exclude zero; red intervals include zero",
       x = "Posterior Mean (Standardized Scale)", y = "", color = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Key Findings

Based on the 95% credible intervals from our best model (`fit_log_age`), the most significant covariates for predicting concrete compressive strength are `cement` and `age` (log-transformed), which have the largest posterior means and intervals that are firmly away from zero. `water` is also significant with a strong negative effect. Covariates such as `coarse_agg` and `fine_agg` have intervals that include zero, indicating weak or no evidence of a meaningful effect on strength — consistent with our earlier frequentist screening. The log-age model was selected as the best formulation based on LOO-CV, as it captures the nonlinear curing process of concrete.

# Predictions on Test Set

```{r}
#| label: Predictions

# prepare test covariates using same log(age) transformation and scaling
X_test_raw <- concrete_test %>% select(-ccs)
X_test_raw$age <- log(X_test_raw$age)

# scale test data using training means and SDs
X_test_scaled <- scale(X_test_raw,
                       center = attr(X_log_age_scaled, "scaled:center"),
                       scale  = attr(X_log_age_scaled, "scaled:scale"))

# extract posterior samples from the best model
alpha_samples <- as.matrix(fit_log_age)[, "alpha"]
beta_samples_pred <- as.matrix(fit_log_age)[, paste0("beta[", 1:8, "]")]

# compute posterior predictive mean for each test observation
# each row of pred_matrix is one posterior draw, each column is one test obs
pred_matrix <- alpha_samples + beta_samples_pred %*% t(X_test_scaled)

# posterior mean predictions (vector of length 230)
post_mean <- colMeans(pred_matrix)

# 95% posterior predictive intervals (230 x 2 matrix)
post_int <- t(apply(pred_matrix, 2, quantile, probs = c(0.025, 0.975)))
colnames(post_int) <- c("2.5%", "97.5%")

# verify dimensions
cat("post_mean length:", length(post_mean), "\n")
cat("post_int dimensions:", dim(post_int), "\n")

# preview
head(data.frame(Mean = round(post_mean, 2), post_int))
```

```{r}
#| label: Export-Results

# save predictions to .RData file
save(post_mean, post_int, file = "40233_Group_3.RData")

```
