---
title: "FinalProject"
author: "Branson Vu, Blake Tran, (insert your name here)"
format: pdf
editor: visual
---

# Exploratory Data Analysis

```{r}
#| label: Setup
#| message: false
#| warning: false

# libraries
library(dplyr)
library(ggplot2)
library(corrplot)
# GGally is a package that creates nice looking scatterplot matrice
library(GGally)
library(tidyr)
#library(car)

# Load the RData file, no need to directly click on the file to load it
load("concrete_data.RData")
```

## Data Overview

```{r}
#| label: Data-Overview

# Dimensions and types
str(concrete_train)

# Summary stats for all variables
summary(concrete_train)
```

```{r}
#| label: Missing-Values-and-Proportion-of-Zeros

# Check for missing values in train set
colSums(is.na(concrete_train))

# Proportion of zeros per covariate (slag, ash, cemet_superplasticizer should have many)
round(colMeans(concrete_train == 0), 3)
```

## Response Variable: ccs

```{r}
#| label: Response-Distribution
#| fig-width: 8
#| fig-height: 4

# Histogram + density of ccs (concrete compressive strength)
ggplot(concrete_train, aes(x = ccs)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  labs(title = "Distribution of CCS", x = "Compressive Strength (MPa)", y = "Density") +
  theme_minimal()
```

```{r}
#| label: Response-Log-Transform
#| fig-width: 8
#| fig-height: 4

# Log(ccs) — check if transformation improves normality
ggplot(concrete_train, aes(x = log(ccs))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  labs(title = "Distribution of log(CCS)", x = "log(Compressive Strength)", y = "Density") +
  theme_minimal()
```

```{r}
#| label: Shapiro-tests

# normality tests on raw and log-transformed ccs
shapiro.test(concrete_train$ccs)
shapiro.test(log(concrete_train$ccs))
```

## Covariate Distributions

```{r}
#| label: Covariate-Histograms
#| fig-width: 10
#| fig-height: 8

# Histograms of all 8 covariates
concrete_train %>%
  select(-ccs) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(title = "Covariate Distributions") +
  theme_minimal()
```

## Correlations

```{r}
#| label: Correlation-Matrix
#| fig-width: 8
#| fig-height: 7

# pearson correlation matrix
cor_mat <- cor(concrete_train)
corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0,0,1,0))
```

```{r}
#| label: Correlations-with-Response

# Correlations with ccs sorted by absolute value
cors <- cor(concrete_train)[, "ccs"]
sort(abs(cors[-length(cors)]), decreasing = TRUE)
```

## Scatterplots vs Response

```{r}
#| label: Scatter-vs-ccs
#| fig-width: 10
#| fig-height: 8

# Scatterplots of each covariate against ccs
concrete_train %>%
  pivot_longer(-ccs, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, y = ccs)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.7) +
  facet_wrap(~variable, scales = "free_x", ncol = 3) +
  labs(title = "Covariates vs CCS", y = "Compressive Strength (MPa)") +
  theme_minimal()
```

Note: Age is likely to be important covariate, maybe Water since most correlate around approximately at 190?

## Multicollinearity Check

```{r}
#| label: Pairs-Top-Covariates
#| fig-width: 10
#| fig-height: 10
#| message: false

# Pairwise plot of top correlated covariates + response
top_vars <- c("cement", "age", "water", "superplasticizer", "ccs")
ggpairs(concrete_train[, top_vars],
        lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5))) +
  theme_minimal()
```

## Quick Header

```{r}
freq_fit <- lm(ccs ~ ., data = concrete_train)
summary(freq_fit)
# check multicollinearity
#car::vif(freq_fit)
```

Age, Cement, Slag, Ash, Water are likely significant covariates, but should include in the model to compare via LOO

# Model Fit and Assessment

```{r}
library(rstan)
#library(rstanarm)
library(loo)
# standardize covariates
X_raw <- concrete_trainselect(-ccs)
X_means <- colMeans(X_raw)
X_sds   <- apply(X_raw, 2, sd)
X_scaled <- scale(X_raw)
```

```{r}
stan_concrete_model <- "
data {
  int<lower=0> N;          // number of observations
  int<lower=0> K;          // number of covariates
  matrix[N, K] X;          // covariate matrix (standardized)
  vector[N] y;             // response (log(ccs) or ccs)
}

parameters {
  real alpha;              // intercept
  vector[K] beta;          // slopes
  real<lower=0> sigma;     // residual sd
}

model {
  // priors
  alpha ~ normal(0, 10);
  beta  ~ normal(0, 5);
  sigma ~ cauchy(0, 5);

  // likelihood
  y ~ normal(alpha + X * beta, sigma);
}

generated quantities {
  vector[N] log_lik;       // for LOO-CV
  for (n in 1:N)
    log_lik[n] = normal_lpdf(y[n] | alpha + X[n] * beta, sigma);
}
"
  
```

```{r}
fit <- stan(stan_concrete_model, data = concrete_train,
            chains = 4, iter = 2000, warmup = 1000, seed = 42)
```
